Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
nested loop example

% All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
% is generate constraints over-approximating %abstracting 
% the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Our technique proves the termination of each loop separately and in program order, 
with the exception of nested loops, whose termination we encode in the same formula.
Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
as the conjunction of constraints on the right side of the figure:
%% =======
%% %
%% As the generation of such constraints is standard and covered by several
%% other
%% works~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08},
%% we will not provide the full algorithm, but rather illustrate it through the
%% example in Figure~\ref{fig:environment-model}.  Let us assume that we have
%% already proven the termination of loops $L_1$ and $L_3$, and we are now
%% trying to prove that $L_2$ terminates, which we encode as the conjunction of
%% constraints on the right-hand side of the figure:
%% %
%% >>>>>>> ee2c21bbc574ab36e7d6739c0e82c7a03d930789
\begin{itemize}
\item $W_1$ and $W_2$ are (over-approximating) inductive invariants for $L_1$ and $L_2$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R_2$ and $R_3$ are ranking functions for $L_2$ and $L_3$, respectively.
\end{itemize}

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
& z' + v' = x' \wedge v' \leq y' \\
R_2(x) & = x\\
R_3 & = ..
\end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.145\textwidth}
\begin{lstlisting}[mathescape=true]
x = 0;
$L_1:$
while(x<=y)
  x++;

$L_2:$
while(x>0){
  z = x;
  v = 0;

$L_3:$
  while(v<y){
    z--;
    v++;
  }

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.85\textwidth}
\begin{align*}
 \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
   S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
   & R_3(v,x,y,z) > 0 \, \wedge \\
   & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R_2(v, x, y, z) > 0 \, \wedge \\
   & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\end{framed}
\caption{A non-trivial program and its termination formula\label{fig:environment-model}}
\end{figure*}

>>>>

% All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
% is generate constraints over-approximating %abstracting 
% the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Our technique proves the termination of each loop separately, 
with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
Figure~\ref{fig:environment-model}:
%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
%% as the conjunction of constraints on the right side of the figure:
\begin{itemize}
\item $T_o$  over-approximates $L_1$'s transition relation.
\item $P$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
\end{itemize}

%% Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
%% Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
%% \begin{align*}
%% W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
%% W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
%% S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
%% & z' + v' = x' \wedge v' \leq y' \\
%% R_2(x) & = x\\
%% R_3 & = ..
%% \end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.16\textwidth}
\begin{lstlisting}[mathescape=true]
$L_1:$
while (i<n){
  j = 0;

$L_2:$
  while (j$\leq$i){
    j = j + 1;
  }

  i = i + 1;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.84\textwidth}
\begin{align*}
 \exists T_o, P, R_1, R_2 . \forall i, j, n, i', j', n' . \\
   %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
   P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow P(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
   & R_2(i,j,n) > 0 \, \wedge \\
   & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
   i < n \wedge T_o(i, j, n) & \rightarrow P(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
   i < n \wedge T_o(i, j, n) \wedge P(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
   & T_o(i'+1, j', n') \, \wedge \\
   & R_1(i, j, n) > 0 \, \wedge \\
   & R_1(i, j, n) > R_1(i'+1, j', n')
\end{align*}
\end{minipage}
\end{framed}
\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
\end{figure*}

%% \begin{figure*}
%% \begin{framed}
%% \begin{minipage}{0.145\textwidth}
%% \begin{lstlisting}[mathescape=true]
%% x = 0;
%% $L_1:$
%% while(x<=y)
%%   x++;

%% $L_2:$
%% while(x>0){
%%   z = x;
%%   v = 0;

%% $L_3:$
%%   while(v<y){
%%     z--;
%%     v++;
%%   }

%%   x -= z;
%%   y -= z;
%% }
%% \end{lstlisting}
%% \end{minipage}
%% \vline
%% \begin{minipage}{0.85\textwidth}
%% \begin{align*}
%%  \exists W_1, W_2, R_2, R_3, S . \forall v, x, y, z, v', x', y', z' . \\
%%    x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
%%    x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
%%    x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
%%    S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' < y' & \rightarrow S(\langle v, x, y, z \rangle, \langle v'+1, x', y', z'-1 \rangle) \, \wedge \\
%%    & R_3(v,x,y,z) > 0 \, \wedge \\
%%    & R_3(v,x,y,z) > R_3(v'+1,x',y',z'-1) \, \wedge\\
%%    x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
%%    x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
%%    & W_2(v', x' - z', y' - z', z') \, \wedge \\
%%    & R_2(v, x, y, z) > 0 \, \wedge \\
%%    & R_2(v, x, y, z) > R_2(v', x' - z', y'-z', z')
%% \end{align*}
%% \end{minipage}
%% \end{framed}
%% \caption{A non-trivial program and its termination formula\label{fig:environment-model}}
%% \end{figure*}
